---
title: "blue whale figures"
author: "DAS"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
require(mosaic)
require(lattice)
require(graphics)
require(ggplot2)
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
knitr::opts_chunk$set(fig.path='figures_bw/', dev=c('pdf', 'png'), dpi=300)

load('bwgg.RData')
```

##Blue Whale Figures
Plot of the averaged blue whale rates for the whales which have known lunge times.
```{r, echo=FALSE, fig.align='center', message=FALSE, comment=FALSE, warning=FALSE}
False_Positive_Rate <- bw_data$rates$avgr.withlunges[2:101,2]
True_Positive_Rate <- bw_data$rates$avgr.withlunges[2:101,1]
avgr <- cbind(True_Positive_Rate, False_Positive_Rate)
ggplot((data.frame(avgr)), aes(x = False_Positive_Rate, y = True_Positive_Rate)) + geom_point() + theme_bw() + theme(axis.text=element_text(size=13), axis.title=element_text(size=15,face="bold")) + labs(x = "False Positive Rate", y = "True Positive Rate") + xlim(0,1) + ylim(0,1)
```

Plot of the averaged blue whale rates for all whales.
```{r, echo=FALSE, fig.align='center', message=FALSE, comment=FALSE, warning=FALSE}
False_Positive_Rate <- bw_data$rates$avgr.all[2:101,2]
True_Positive_Rate <- bw_data$rates$avgr.all[2:101,1]
avgr <- cbind(True_Positive_Rate, False_Positive_Rate)
ggplot((data.frame(avgr)), aes(x = False_Positive_Rate, y = True_Positive_Rate)) + geom_point() + theme_bw() + theme(axis.text=element_text(size=13), axis.title=element_text(size=15,face="bold")) + labs(x = "False Positive Rate", y = "True Positive Rate") + xlim(0,1) + ylim(0,1)
```
\clearpage

Plot of the medians of the blue whale rates for the whales which have known lunge times.
```{r, echo=FALSE, fig.align='center', message=FALSE, comment=FALSE, warning=FALSE}
False_Positive_Rate <- bw_data$rates$med.withlunges[2:101,2]
True_Positive_Rate <- bw_data$rates$med.withlunges[2:101,1]
med <- cbind(True_Positive_Rate, False_Positive_Rate)
ggplot((data.frame(med)), aes(x = False_Positive_Rate, y = True_Positive_Rate)) + geom_point() + theme_bw() + theme(axis.text=element_text(size=13), axis.title=element_text(size=15,face="bold")) + labs(x = "False Positive Rate", y = "True Positive Rate") + xlim(0,1) + ylim(0,1)
```

Plot of the medians of the blue whale rates for all whales.
```{r, echo=FALSE, fig.align='center', message=FALSE, comment=FALSE, warning=FALSE}
False_Positive_Rate <- bw_data$rates$med.all[2:101,2]
True_Positive_Rate <- bw_data$rates$med.all[2:101,1]
med <- cbind(True_Positive_Rate, False_Positive_Rate)
ggplot((data.frame(med)), aes(x = False_Positive_Rate, y = True_Positive_Rate)) + geom_point() + theme_bw() + theme(axis.text=element_text(size=13), axis.title=element_text(size=15,face="bold")) + labs(x = "False Positive Rate", y = "True Positive Rate") + xlim(0,1) + ylim(0,1)
```
\clearpage

Plots of individual whales in numerical order.
```{r}
require(ggformula)
FP <- list()
TP <- list()
r <- list()
for (k in 5:(length(bw_data$rates))) {
  FP[[k]] <- bw_data$rates[[k]][2:101,1]
  TP[[k]] <- bw_data$rates[[k]][2:101,2]
  r[[k]] <- cbind(TP[[k]], FP[[k]])
  gf_point(r[[k]][,2]~r[[k]][,1], col="black") %>% 
    gf_labs(title='Blue Whale Detections', x='False Positive Rate', y='True Positive Rate') %>% 
    gf_lims(x=c(0,1), y=c(0,1)) %>% 
    gf_theme(theme_minimal())
}
```

```{r, echo=FALSE, fig.align='center', message=FALSE, comment=FALSE, warning=FALSE, fi.height=20, fig.width=7}
for (j in c(1:14,16:length(bw_data$fs))) {
  ggplott(bw_data$A[[j]],bw_data$fs[[j]],bw_data$p[[j]],bw_data$lunge[[j]],bw_table$`Optimal Threshold`[j],30,n=2,species='whale')
}
```