---
title: "Code folding and hiding figures and other output"
author: "tagtools project team"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{code-and-figure-folding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hiding a figure/output

Since the YAML header (at the top of the document) has `code_folding: hide` there is a little button to toggle showing/hiding the code.

Since `eval = FALSE` below, the code will be shown but will not be run.


```{r cars-figure, eval = FALSE}
plot(mtcars$hp, mtcars$mpg)
```

Now add a little button for "showing results" (unfortunately this requires having a second "copy" of the chunk and the code to make the figure or do the calculation).  For the button, add the html code below BEFORE the R chunk. Change the block name "car_figure" so it matches in the "button" and the "div", and is unique for each instance:

<button class="btn btn-primary" data-toggle="collapse" data-target="#car_figure"> Show/Hide Results </button>  
<div id="car_figure" class="collapse">  

```{r cars-figure-just-to-show-results, echo = FALSE, eval = TRUE}
plot(mtcars$hp, mtcars$mpg)
```

</div>

(don't forget that end div after the code chunk!)

