---
title: "Using dive_stats"
author: "tagtools project team"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{dive-stats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tagtools)
```

Welcome to this vignette! On behalf of the team behind tagtools, thanks for taking some time to get to know our package. We hope it sparks joy.

In this vignette you will learn to use the function `dive-stats` to compute summary statistics for a given dataset containing dives, including adding an auxiliary variable.

*Estimated time for this vignette: 25 minutes.*

These practicals all assume that you have R/Rstudio, some basic experience working with them, and can execute provided code, making some user-specific changes along the way (e.g. to help R find a file you downloaded). We will provide you with quite a few lines. To boost your own learning, you would do well to try and write them before opening what we give, using this just to check your work.

Additionally, be careful when copy-pasting special characters such as \_underscores\_ and 'quotes'. If you get an error, one thing to check is that you have just a single, simple underscore, and `'straight quotes'`, whether `'single'` or `"double"` (rather than “smart quotes”). 

# Finding dives and summarising them

Consider a dataset from a DTAG attached to a Cuvier’s beaked whale, Ziphius cavirostris. Load the data from file zc11_267a.nc. This dataset is built into the `tagtools` package, so you can access it using `system.file`.

```{r, echo = TRUE, eval = FALSE}
library(tagtools)
ZC_file_path <- system.file("extdata", "zc11_267a.nc", package = "tagtools", mustWork = TRUE)
ZC <- load_nc(ZC_file_path)
```

```{r, eval = TRUE, echo = FALSE}
ZC_file_path <- system.file("extdata", "zc11_267a.nc", package = "tagtools", mustWork = TRUE)
ZC <- load_nc(ZC_file_path)
```

1. Make a plot of the dive profile. What do you notice?

```{r, echo = TRUE, eval = FALSE}
plott(X=list(Depth=ZC$P), r = TRUE)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#plott_zc"> Show/Hide Results/*Answers* </button>  
<div id="plott_zc" class="collapse"> 

```{r, echo = FALSE, eval = TRUE}
plott(X=list(Depth=ZC$P), r = TRUE)
```

*There are some very deep dives in this profile! This is perhaps to be expected for Cuvier's beaked whales, with their reputation for mysteriously extreme excursions.*

</div>

2. You probably want to crop the data before further analysis, because there is a period at the start of the recording when the tag was not yet deployed on the whale.

```{r, echo = TRUE, eval = FALSE}
ZPCr = crop(ZC$P)
# if you want to review history/other fields
str(ZPCr, max.level = 1)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#crop_zc"> Show/Hide Results </button>  
<div id="crop_zc" class="collapse"> 
```{r, echo = FALSE, eval = TRUE}
print("Results for `crop(ZC$P)`:")
print("---------------------------------------")
ZPCr = crop(ZC$P)
# if you want to review history/other fields
print("Results for `str(ZPCr, max.level = 1)`:")
print("---------------------------------------")
str(ZPCr, max.level = 1)
```
</div>

```{r, echo = FALSE, eval = TRUE}
ZPCr = crop_to(ZC$P, sampling_rate = ZC$P$sampling_rate, tcues = c(9424, 64027), T = NULL)
```

3. What minimum depth threshold do you think you would use to detect dives this animal's dives? Consider how you would justify your choice.

<button class="btn btn-primary" data-toggle="collapse" data-target="#ans_depth_thresh"> *Show/Hide Answers* </button>  
<div id="ans_depth_thresh" class="collapse"> 
*No less than fifty meters seems like a reasonable choice to detect this animal's dives. After all, most of the dives this animal performs are more than 100 meters in depth.*
</div>

4. Use `find_dives` to detect all dives deeper than your chosen minimum depth `mindepth`.

```{r, echo = TRUE, eval = FALSE}
mindepth <- # your chosen minimum depth here
dt <- find_dives(ZPCr, mindepth=mindepth) 
```

```{r, echo = FALSE, eval = TRUE}
mindepth <- 50 
dt <- find_dives(ZPCr, mindepth=mindepth) 
```

5. Now use `dive_stats` to compute summary statistics for all the dives
you detected. 

```{r, echo = TRUE, eval = FALSE}
ds <- dive_stats(ZPCr, dive_cues=dt[,c('start', 'end'),]) 
str(ds)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#str_ds"> Show/Hide Results </button>  
<div id="str_ds" class="collapse"> 
```{r, echo = FALSE, eval = TRUE}
ds <- dive_stats(ZPCr, dive_cues=dt[,c('start', 'end'),]) 
str(ds)
```
</div>

6. Have a look at the dive stats and perhaps make a plot of some or all
of them. Do you notice anything interesting? Code for a basic example plot (plotting the max depth of each against the to_rate of each dive) is shown below.

```{r, echo = TRUE, eval = FALSE}
plot(x = ds$max, y = ds$from_rate)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#ds_basic_plot"> Show/Hide Results/*Answers* </button>  
<div id="ds_basic_plot" class="collapse"> 

```{r, echo = FALSE, eval = TRUE}
plot(x = ds$max, y = ds$to_rate)
```

*There appears to be little, if any, correlation between the to-rate and the maximum depth. ¯\\_(ツ)_/¯ *

</div>

## Optional: choose an extra variable

7. Choose an auxiliary variable (could be anything of interest - pitch, roll, heading, MSA, ODBA, njerk...). 

Compute the auxiliary variable, and then recompute the dive stats including the auxiliary variable. 

*One example auxiliary variable---pitch---is given below.*

### Compute pitch

The function `a2pr` will calculate pitch and roll from acceleration. To use the acceleration data, we'll want it cropped to the same time window as we've cropped the depth data. 

```{r, echo = TRUE, eval = FALSE}
ZACr <- crop_to(ZC$A, tcues = ZPCr$crop)
# plot to confirm the cropping

plott(X=list(CroppedDepth=ZPCr, CroppedAccel=ZACr), 
      r=c(TRUE,FALSE)) # and reverse the y-axis for depth, but not acceleration 
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#plott_ZPCr_ZACr"> Show/Hide Results </button>  
<div id="plott_ZPCr_ZACr" class="collapse"> 
```{r, echo = FALSE, eval = TRUE}
ZACr <- crop_to(ZC$A, tcues = ZPCr$crop)
# plot to confirm the cropping

plott(X=list(CroppedDepth=ZPCr, CroppedAccel=ZACr), 
      r=c(TRUE,FALSE)) # and reverse the y-axis for depth, but not acceleration 
```
</div>

And now to compute pitch:

```{r, echo = TRUE, eval = TRUE}
ZPiCr <- a2pr(ZACr)$p # this is the variable you'll want to use later
```

### Recompute dive stats

Use the auxiliary variable of your choice (if you were following the example above, you'll use ZPiCr) for my_aux_var.

```{r, echo = TRUE, eval = FALSE}
dsAux <- dive_stats(P = matrix(ZPCr$data), sampling_rate = ZPCr$sampling_rate, dive_cues=dt[,c('start', 'end'),], X=matrix(my_aux_var))
# if you are using ZPiCr
dsAux <- dive_stats(P = matrix(ZPCr$data), 
                    dive_cues=dt[,c('start', 'end'),], 
                    X = matrix(ZPiCr), # to avoid nrow != ncol errors
                    angular = TRUE, # because this is pitch angle data
                    sampling_rate = c(5, 5) # sampling rate is 5 for P & X
                    )
```

```{r, echo = FALSE, eval = TRUE}
dsAux <- dive_stats(P = matrix(ZPCr$data), 
                    dive_cues=dt[,c('start', 'end'),], 
                    X = matrix(ZPiCr), # to avoid nrow != ncol errors
                    angular = TRUE, # because this is pitch angle data
                    sampling_rate = c(5, 5) # sampling rate is 5 for P & X
                    )
```

8. Examine and/or plot again.

```{r, echo = TRUE, eval = FALSE}
str(ds, max.level = 1)
str(dsAux, max.level = 1) 
plott( # ... figure out what you want to plott; results not shown for this part
  )

```

<button class="btn btn-primary" data-toggle="collapse" data-target="#str_ds_dsAux"> Show/Hide Results </button>  
<div id="str_ds_dsAux" class="collapse"> 

```{r, echo = FALSE, eval = TRUE}
print("Results for `str(ds, max.level = 1):`")
print("-----------------------------------------")
str(ds, max.level = 1)
print("Results for `str(dsAux, max.level = 1):`")
print("-----------------------------------------")
str(dsAux, max.level = 1) 
```

</div>

# Review 

You've learned a couple of ways to use `dive-stats`. Well done!

*If you'd like to continue working through these practicals, consider `rotation-test` or `fine-scale-tracking`.*

***

Animaltags home pages: http://animaltags.org/ (old), https://animaltags.netlify.app/ (new), https://github.com/stacyderuiter/TagTools (for source code)