---
title: "dive-stats"
author: "tagtools project team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dive-stats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to this vignette! On behalf of the team behind tagtools, thanks for taking some time to get to know our package. We hope it is useful to you.

In this vignette you will learn to use the function `dive-stats`.

# Finding dives and summarising them

Now consider a dataset from a DTAG attached to a Cuvier’s beaked whale, Ziphius cavirostris. Load the data from file zc11_267a.nc. (As usual: Make sure that `zc11_267a.nc` is in your working directory. Is `load_nc()` not in the mood to exist? Then `library(tagtools)` is your friend. How about if `zc11_267a.nc` decides to actually be `no such file or directory`? Well, search for it in your file finder. After that, `getwd()` and `setwd("PathToWorkingDirectoryInQuotes")` are your friends.)  


```{r, eval = FALSE}
ZC <- load_nc('zc11_267a')
```

1. Make a plot of the dive profile. What do you notice?

```{r, eval = FALSE}
plott(X=list(Depth=ZC$P), r = TRUE)
```

2. You probably want to crop the data before further analysis, because there is a period at the start of the recording when the tag was not yet deployed on the whale.

```{r, eval = FALSE}
Zpc = crop(ZC$P)
# if you want to review history/other fields
str(Zpc, max.level = 1)
```

3. What minimum depth threshold do you think you would use to detect dives by this animal? Consider how you would justify your choice.

4. Use find dives to detect all dives deeper than your chosen depth min- depth.

```{r}
dt <- find_dives(ZC$P, mindepth=mindepth) 
```

5. Now use dive stats to compute summary statistics for all the dives
you detected. 

```{r, eval = FALSE}
ds <- dive_stats(ZC$P, dive_cues=dt[,c('start', 'end'),]) 
```

6. Have a look at the dive stats and perhaps make a plot of some or all
of them. Do you notice anything interesting?

## Optional: choose an extra variable

7. Choose an auxiliary variable (could be anything of interest - pitch, roll, heading, MSA, ODBA, njerk...). 

Compute the auxiliary variable, and then recompute the dive stats including the auxiliary variable. 

<!-- Example auxiliary variable: pitch. More content to come next week. -->

```{r, eval = FALSE}
ds <- dive_stats(zc$P, dive_cues=dt[,c('start', 'end'),],
                 X=my_aux_var)
```

8. Examine and/or plot again.

```{r, eval = FALSE}
str(ds, max.level = 1)
plott()

```

# Review 

You've learned a couple of ways to use `dive-stats`. Well done!