---
title: "Setting up tagtools"
author: "tagtools project team"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{install-load-tagtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE
)
```

Welcome to this vignette! On behalf of the team behind tagtools, thanks for taking some time to get to know this package. We hope it is useful to you. 

In this vignette you will complete the first important setup steps of installing the package on your machine and making sure it is loaded in your session.

*Estimated time for this vignette: 15 minutes*

# Installing tagtools for R

## Prerequisites
Before you begin, you must have R and---recommended---RStudio installed on your computer. You can download this software at [RStudio.com](https://www.rstudio.com/products/rstudio/#rstudio-desktop) (you only need the free one). Alternatively, you *can* work in [RStudio.cloud](https://rstudio.cloud). This is free too, and it is nice that no download is required. The drawback is that, with your free RStudio.cloud account, you are limited at a certain number of project-hours per month, RAM, and CPU. One who does a lot of work in RStudio, several hours a day, will run out of hours within a week or two. The desktop version has fewer restraints. 

If you happen to be new to R also, here are two very nice standalone tutorials: [#1](https://rsconnect.calvin.edu:3939/content/141) and [#2](https://rstudio.cloud/learn/primers/1.2).

These practicals assume that you have some basic experience working with R/RStudio, and can execute provided code, making some user-specific changes along the way (e.g. to help R find a file you downloaded). We will provide you with quite a few lines. To boost your own learning, you would do well to try and write them before opening what we give, using this just to check your work. 

## Install tagtools from github
If you have `devtools` installed on your computer, you can use it to install the `tagtools` package directly from github.

If you want or need to install `devtools`, or want more information about it, see the [online installation guide at r-project.org](https://www.r-project.org/nosvn/pandoc/devtools.html).

The code below gives the devtools command to install. Click "Code" at the right to open the chunk, then you should be able to copy-paste the code into your R console. **However**, be careful when copy-pasting special characters such as \_underscores\_, and 'quotes'. If you get an error, one thing to check is that you have just a single, simple underscore, and `'straight quotes'`, whether `'single'` or `"double"` (rather than “smart quotes”). 

```{r, eval = FALSE}
devtools::install_github('stacyderuiter/TagTools/R/tagtools',
                         build_vignettes = TRUE)
```

If you are asked soomething like "These packages have more recent versions available", it is no harm to update them all, and to "install from sources the packages which need compilation". 

With the package as of July 2021, the overall process should take ten to fifteen minutes, depending on the speed of your internet and machine, some eight minutes of which are spent building vignettes. 

## Install tagtools from zip archive
If you do not want to use the `devtools` option, you can install the `tagtools` package from the archive files provided with your workshop materials (in the folder created by unzipping wmmc-2019-workshop-materials.zip). These archive files are also available at the TagTools website. Try http://animaltags.org , or https://animaltags.netlify.app . 

*Note: if you choose to install from archive files (.zip or .tgz), you will probably also have to manually install all the packages `tagtools` depends on. If you use the `devtools` method above, you will not.*

- **Windows** users will want to use the **.zip** archive
- **Mac** or **Linux** users will want to use the **.tgz** archive

Save the appropriate archive to a known location on your computer.

Then, run:

```{r, eval = FALSE}
install.packages('YourPath/YourFilename') # you'll have to edit this appropriately
```

(fill in the absolute or relative path to the archive file, and the archive file name). For example, on a Mac, if the file is saved to Downloads, your path might look like this: 

```{r, eval = FALSE}
install.packages('/Users/YourUsernameHere/Downloads/FileName.tgz')
```

You can also use the "install packages" GUI in RStudio---go to the "Packages" tab, click "Install", choose "Package Archive File" from the "Install From" pulldown menu, and then navigate to the archive file. Then follow the prompts.

You must also make sure you have installed all the packages that `tagtools` depends on:

```{r, eval = FALSE}
dpnds <- c('CircStats', 'ggformula', 'graphics', 'hht', 
           'latex2exp', 'lubridate', 'magrittr', 
           'matlab', 'ncdf4', 'plotly', 'pracma', 
           'readr', 'rgl', 'signal', 'stats', 
           'utils', 'zoo', 'zoom')
install.packages(pkgs = dpnds)
```

Note here that we are writing an object `dpnds`, which is a vector of these 18 (text) strings, concatenated together with `c()`. Then `install.packages()` can take all these dependencies as an input.

## Check tagtools Installation

Once the `tagtools` are successfully installed, you should be able to run

```{r, eval = FALSE, echo = TRUE}
library(tagtools)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#library_tagtools"> Show/Hide Results </button>  
<div id="library_tagtools" class="collapse"> 

```{r, eval = TRUE, echo = FALSE}
library(tagtools)
```

</div>

You will often get this message: "The following object is masked from 'package:stats': smooth"---this means you have loaded it properly; congratulations! 

Another check is to make sure that function help files are accessible; for example,

```{r, echo = TRUE, eval = FALSE}
?load_nc
```

Now or at any point in the future when doing a check like this, you might get an error saying the function doesn't exist. When this happens, make sure you have tagtools loaded in your library. While you won't have to reinstall the package with `install.packages()` unless you start working on a different machine, you will have to run `library(tagtools)` or `require(tagtools)` every time you restart RStudio. 

Having fewer packages loaded is kinder to your computer's memory, so it is good that R doesn't automatically load everything. However, it does mean that every time you restart RStudio, you must reload the packages you need, such as `tagtools`, yourself. Thus at the beginning of each vignette after this one you'll be reminded to run `library(tagtools)`. 

If the command `? load_nc` pulls up a help file for the function `load_nc` like it should, you're all set. 

# Accessing other vignettes through R directly

If you've run the `devtools` command, you can read these vignettes directly in your local RStudio without needing to click through the website! The syntax is very straightforward. To see the list of vignettes in just this package, simply type: 

```{r, echo = TRUE, eval = FALSE}
vignette(package = "tagtools")
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#vignettes_in_tagtools"> Show/Hide Results </button>  
<div id="vignettes_in_tagtools" class="collapse"> 
```{r, echo = FALSE, eval = TRUE}
print("Vignettes in package ‘tagtools’:")
print("")
print("complementary-filtering         complementary-filtering (source, html)")
print("data-quality-error-correction   data-quality-error-correction (source, html)")
print("Detectors                       Detectors (source, html)")
print("detectors-draft                 detectors-draft (source, html)")
print("dive-stats                      dive-stats (source, html)")
print("find-dives                      find-dives (source, html)")
print("fine-scale-tracking             fine-scale-tracking (source, html)")
print("install-load-tagtools           install-load-tagtools (source, html)")
print("jerk-transients                 jerk-transients (source, html)")
print("load-tag-data                   load-tag-data (source, html)")
print("mahalanobis-distance            mahalanobis-distance (source, html)")
print("more-filtering                  more-filtering (source, html)")
print("plots-and-cropping              plots-and-cropping (source, html)")
print("rotation-test                   rotation-test (source, html)")
print("tag-to-whale-frame              tag-to-whale-frame (source, html)")
print("vectors-vs-structures           vectors-vs-structures (source, html)")
```
</div>

Then you can run any vignette with `vignette('package-index', package = 'tagtools')`. For instance, in order to see this vignette, you can run:

```{r, echo = TRUE, eval = FALSE}
vignette("install-load-tagtools", package = "tagtools")
```

The `package = "tagtools"` argument at the end is optional. However, it is a good addition because it clarifies which package you are working in. So, if there happened to be, say, another `vectors-vs-structures` vignette in some other package, you could load the one that is specific to this package with `vignette("vectors-vs-structures")`. 

# Review

What have you learned so far? Important setup steps that will help avoid confusion and delay moving forward.

Congrats!

*If you'd like to continue working through these vignettes, `load-tag-data` and `plots-and-cropping` are good options to help you get started with the tools you've just installed and loaded.*

```{r, echo = TRUE, eval = FALSE}
vignette('load-tag-data', package = 'tagtools')
vignette('plots-and-cropping', package = 'tagtools')
```

***

Animaltags home pages: http://animaltags.org/ (old), https://animaltags.netlify.app/ (new), https://github.com/stacyderuiter/TagTools (for source code)