---
title: "Jerk: a way to emphasize rapid changes in acceleration"
author: "tagtools project team"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{jerk-transients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
require(tagtools)
```

Welcome to the jerk-transients vignette! Thanks for getting some time to know our package. We hope it's, thus far, all you've dreamed it would be.

In this vignette, you will use jerk to identify prey capture attempts. 

*Estimated time for this vignette: 20 minutes.*

These practicals all assume that you have R/Rstudio installed on your machine, some basic experience working with them, and can execute provided code, making some user-specific changes along the way (e.g. to help R find a file you downloaded). We will provide you with quite a few lines. To boost your own learning, you would do well to try and write them before opening what we give, using this just to check your work.

Additionally, be careful when copy-pasting special characters such as \_underscores\_ and 'quotes'. If you get an error, one thing to check is that you have just a single, simple underscore, and `'straight quotes'`, whether `'single'` or `"double"` (rather than “smart quotes”). 

# Setup
If you've finished `complementary-filtering` within the same R session, skip to [Acceleration transients, jerk and dynamic acceleration]. If not, you'll want to do this first section. 

For this practical we will use data from a suction cup tag attached to the back of a beaked whale. This dataset is built into the `tagtools` package, so you can access it using `system.file`.  

Write `testset1.nc` to the object `bw`, for "beaked whale". Then use `plott()` to get a look at it.

```{r, echo = TRUE, eval = FALSE}
library(tagtools)
bw_file_path <- system.file("extdata", "testset1.nc", package = 'tagtools', mustWork = TRUE)
bw <- load_nc(bw_file_path)
plott(X = list(Depth = bw$P, Acc = bw$A, Mag = bw$M))
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#plott_bw"> Show/Hide Results </button>  
<div id="plott_bw" class="collapse">

```{r, echo = FALSE, eval = TRUE}
bw_file_path <- system.file("extdata", "testset1.nc", package = 'tagtools', mustWork = TRUE)
bw <- load_nc(bw_file_path)
plott(X = list(Depth = bw$P, Acc = bw$A, Mag = bw$M))
```

</div>

This dataset contains a deep dive followed by a shallow dive. We want to infer the function of these by looking for locomotion effort and sudden changes in acceleration that could be indicative of prey capture attempts. We are also going to look for changes in swimming gait.

# Acceleration transients, jerk and dynamic acceleration
We want to look for indications of foraging during the two dives. Sudden changes (transients) in acceleration are often associated with prey capture attempts. One way to emphasise rapid changes in acceleration is by differentiating it to produce the jerk. 

This is effectively high-pass filtering acceleration with a constantly-sloping filter---the higher the frequency, the more it is emphasised. Because we don't know what kind of movement is involved in a prey capture attempt, we don't know which axis of acceleration will be most relevant. It is safer then to just compute the magnitude of the jerk in all three axes to get a single vector (also called "norm-jerk"). That way, a peak in any or all axes will show up. `njerk` computes this from acceleration. Try running `njerk` on our acceleration data (where is it stored?) and writing the result to `J`. Then, once you also grab the sampling rate, you can plot this jerk data. 

```{r, eval = FALSE, echo = TRUE}
J <- njerk(bw$A)
fs <- bw$A$sampling_rate
plott(X = list(Depth = bw$P$data, Jerk = J), fsx = fs)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#plott_J"> Show/Hide Results </button>  
<div id="plott_J" class="collapse">

```{r, eval = TRUE, echo = FALSE}
J <- njerk(bw$A)
fs <- bw$A$sampling_rate
plott(X = list(Depth = bw$P$data, Jerk = J), fsx = fs)
```

</div>

Where in the dive profile do the largest jerk transients (peaks) appear? Check out some of the biggest peaks to see how short they are.

## Effects of Sampling Rate on jerk
The size and clarity of the jerk peaks depends on the sampling rate. We are using 25 Hz data. To see what you would get with 5 Hz data, try decimating the acceleration before computing the jerk:

```{r, echo = TRUE, eval = FALSE}
str(bw$A, max.level = 1) # look at the data before decimation
Ad <- decdc(bw$A, 5) # decimate by 5 times. New sampling rate is 25/5 Hz
str(Ad, max.level = 1) # and after decimation
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#Ad"> Show/Hide Results </button>  
<div id="Ad" class="collapse">

```{r, echo = FALSE, eval = TRUE}
print("results for str(bw$A, max.level = 1)")
print("------------------------------------")
str(bw$A, max.level = 1)
Ad <- decdc(bw$A, 5) # decimate by 5 times. New sampling rate is 25/5 Hz
print("results for str(Ad, max.level = 1)")
print("------------------------------------")
str(Ad, max.level = 1)
```

</div>

Compute the decimated jerk from the decimated acceleration, and plot it with the jerk computed from the full bandwidth acceleration:

```{r, echo = TRUE, eval = FALSE}
Jd <- njerk(Ad)
plott(X = list(Depth = bw$P$data, `25 Hz Jerk` = J, `5Hz Jerk` = Jd),
      fsx = c(fs, fs, fs/5))
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#plot_jd"> Show/Hide Results </button>  
<div id="plot_jd" class="collapse">
```{r, echo = FALSE, eval = TRUE}
Jd <- njerk(Ad)
plott(X = list(Depth = bw$P$data, `25 Hz Jerk` = J, `5Hz Jerk` = Jd),
      fsx = c(fs, fs, fs/5))
```
</div>

Check out some of the obvious peaks around minute 20 to minute 30 to have a look at the 'signal-to-noise' in the jerk transients. How clear are the jerk transients in the decimated signal compared to the original data rate? What does this tell you about the frequency content in the transients?

<!-- AGHH I need to understand this.  -->

## Detecting Jerk Peaks
To find potential prey capture attempts, we need to run a detector on the norm-jerk signal. Although it is easy to see peaks by eye in the data, detectors require some information to do a good job: they need to know the threshold above which a peak is really a peak and the blanking time, i.e., the minimum time that must elapse after a detection before another detection can happen. 

The tagtools includes interactive peak detectors that allow you to choose a threshold and see the effect this has on which transients are detected. We will use a blanking time of 5s, i.e., we expect that the shortest time between prey captures is 5s:

```{r, echo = TRUE, eval = FALSE}
pks <- detect_peaks(J, fs, bktime = 5)	# 5s blanking time, threshold is interactive
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#detect_peaks"> Show/Hide Results </button>  
<div id="detect_peaks" class="collapse">

```{r, echo = FALSE, eval = TRUE}
pks <- detect_peaks(J, fs, bktime = 5)	# 5s blanking time, threshold is interactive
```

</div>

Follow the instructions in the console to change the threshold. You need to pick a threshold that excludes most of the jerk transients during the strong propulsion locomotion in the ascent but that still detects most of the jerk transients during the bottom part of the deep dive. (Balancing false detections and missed detections is often not easy in a detector and is a matter of finding a trade-off that works for your application.)

This function returns a list of information about the detected peaks. For each detection, the start and end times are reported (in seconds), along with the time at which the peak occurred and the height of the peak. Finally, the selected threshold and blanking time are reported. You can plot the height and time of each detection along with the dive profile as follows:

```{r, echo = TRUE, eval = FALSE}
plott(X = list(Depth = bw$P))
# note: the code below assumes your plott x-axis is in hours.
# if it were in minutes use /60 instead of /3600, etc.
points(pks$peak_time/fs/3600, bw$P$data[round(pks$peak_time)], pch = 8)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#plot_peaks"> Show/Hide Results </button>  
<div id="plot_peaks" class="collapse">

```{r, echo = FALSE, eval = TRUE}
plott(X = list(Depth = bw$P))
# note: the code below assumes your plott x-axis is in hours.
# if it were in minutes use /60 instead of /3600, etc.
points(pks$peak_time/fs/3600, bw$P$data[round(pks$peak_time)], pch = 8)
```

</div>

Bearing in mind that some of the jerk peaks might come from strong locomotory strokes, is there strong evidence for foraging in the deep dive? Is there strong evidence for foraging in the shallow dive?

<button class="btn btn-primary" data-toggle="collapse" data-target="#ans_foraging_shallow_dive"> _Show/Hide Answers_ </button>  
<div id="ans_foraging_shallow_dive" class="collapse">
*Answer text...*
</div>

Now, if you have done `complementary-filtering` recently, here is one more question for you. Adding your inference from `complementary-filtering` about locomotion in the shallow dive, what can you conclude about the behaviour of the animal in the shallow dive? Is it resting, traveling or foraging? Write down your marvelous thoughts.

<button class="btn btn-primary" data-toggle="collapse" data-target="#ans_behav_shallow_dive"> _Show/Hide Answers_ </button>  
<div id="ans_behav_shallow_dive" class="collapse">
*Answer text...*
</div>

# Review

You've learned how to accentuate quick movements of an animal using jerk, and used the accentuated picture to gain understanding about the animal's behaviour. 

And with that, congrats! You bounced through this vignette. 

*If you'd like to continue working through these vignettes, `complementary-filtering` and `more-filtering` are very logical choices. You'll gain more insight into the same dataset, and other ways of analysing locomotion, by doing these.*

***

Animaltags home pages: http://animaltags.org/ (old), https://animaltags.netlify.app/ (new), https://github.com/stacyderuiter/TagTools (for source code)